---
title: 'Day1: Data visualization basics'
author: "Wenbin Guo"
date: "2024-02-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Hello everyone, welcome to the QCBio workshop W28: `Advanced Data Visualization with ggplot2`

# Beginning Remarks
In today’s fast-paced research environment, the ability to communicate complex data effectively is invaluable. ggplot2, a powerful package in R, stands out for its versatility and capacity to create aesthetically pleasing and complex visualizations with ease. This workshop is designed to help you enhance data visualization skills using ggplot2 and enabling you to craft and tailor plots that effectively communicate your research findings.

## Workshop Goals

Our goals for this workshop are threefold:

1. **Deepen Understanding and Skills with ggplot2:** In this workshop, we aim to provide you a deep understanding of ggplot2's grammar and syntax, empowering you to leverage its full potential to create visually appealing and informative graphics.
2. **Address Data Wrangling and Visualization Challenges:** Through practical exercise, we’ll tackle some data wrangling and visualization challenges. We hope the expertise you gained in this workshop can be readily applied to the unique challenges presented by your research.
3. **Learn Principles of Effective Data Visualization:** We'll explore critical data visualization principles and strategies to evade typical errors, providing you with the tools to showcase your data compellingly and attractively.


Upon completing this workshop, you should be well-equipped to present your data in a more comprehensible and compelling way. Let's embark on this visual exploration journey together and discover how to bring your data to life in new and exciting ways!


## Day1 outline
For today's workshop practice, we will cover the following topics
- Environment setup (install packages)
- Getting started with `ggplot()`
- Data wrangling skills
- Aesthetic mappings


## Tips on Rmarkdown usage in Rstudio:
- Ctrl/Cmd + Alt + i: insert a code chunk
- Ctrl/Cmd + Shift + c: comment/uncomment the code

For more information about the Rmarkdown usage, please refer to the following reference:
- [cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf)
- [R markdown cookbook](https://bookdown.org/yihui/rmarkdown-cookbook/)
- [keyboard shortcuts](https://bookdown.org/yihui/rmarkdown-cookbook/rstudio-shortcuts.html)


# Environment setup
install and load the necessary package
```{r}
# install/load package
if(!require(tidyverse)){
  install.packages("tidyverse")
}

library(tidyverse)
```


# Getting started with ggplot2
## load data mpg
Fuel economy data from 1999 to 2008 for 38 popular models of cars. 
```{r}
# load mpg
data(mpg)

# check the first several lines
head(mpg)
```

meaning of variables:
- `cty` and `hwy` record miles per gallon (mpg) for city and highway driving.
- `displ` is the engine displacement in liters.
- `drv` is the drivetrain: front wheel (f), rear wheel (r) or four wheel (a).
- `model` is the model of car. 
- `class` is a categorical variable describing the type of car: two seater, SUV, compact, etc.



## build basic ggplot object
```{r}
base = ggplot(mpg, aes(hwy, displ))
base
```

we can use `$` to check what's inside the base (ggplot object)
```{r}
print(names(base))
```


## modify plots iteratively
```{r}
base + 
  geom_point()
```

```{r}
base + 
  geom_point(aes(colour = drv)) 
```

```{r}
base + 
  geom_point(aes(colour = drv)) + 
  geom_smooth(method = "lm") 
```


```{r}
base + 
  geom_point(aes(colour = drv)) + 
  geom_smooth(method = "lm") + 
  theme_minimal() 
```

```{r}
base + 
  geom_point(aes(colour = drv)) + 
  geom_smooth(method = "lm") + 
  theme_minimal() + 
  ggtitle("Relationship between variables in mpg dataset", subtitle = "hwy vs displ")
```

note: to add the black border line around the plot, add:
> + theme(plot.background = element_rect(color = "black", linewidth = 1.5))


# Data wrangling basics
## variables
### variable types
```{r}
x = c(0, 1.5, -2)
y = c("low", "medium", "high")
```

### factors
check the results of following 2 lines about factors, what do you find?
```{r}
y1 = factor(y)
y2 = factor(y, levels = c("low", "medium", "high"))
```


## data frame
the `mpg` dataset is a typical example of data frame

### create a new data frame
```{r}
df = data.frame(x = c(1,2,3), y = c("low", "medium", "high"))
df2= data.frame(x = c(1,2,3), y0 = c("apple", "banana", "organge"))

head(df)
```

## Manipulate observations
### filter
Filtering is useful for extracting outliers or remove abnormal data points (such as missing values). 

Filtering is also a useful trick when we dealing with big datasets. By filtering, we can explore a small part of the data before we scale up to the whole dataset. This could save us time when building the prototype of processing and visualization pipeline 

```{r}
# singlecomparison operators
filter(mpg, model == "a4")

# with logical operators (compound condition)
filter(mpg, model == "a4" & cyl==4)
```


### arrange
```{r}
# default ascending
arrange(mpg, hwy)

# descending
arrange(mpg, desc(hwy))
```


### merge by rows
use `bind_rows()` to merge 2 dataframe by rows
```{r}
mpg_df1 = filter(mpg, manufacturer == "audi")
mpg_df2 = filter(mpg, manufacturer == "ford")

bind_rows(mpg_df1, mpg_df2)
```


**Exercise**: extract observations whose `cty` value is among the top 10

**Exercise**: extract observations whose manufacturer is `audi` or `ford`, filter by `cty > 18` , and then arrange the resulting data frame by `cty` in descending order

**Exercise**: run `bind_rows(df,df2)`, what would you find?


## Manipulate variables
### select
```{r}
mpg_df3 = select(mpg, manufacturer, model, year, displ)
mpg_df4 = select(mpg, manufacturer, model, year, hwy, cty)

mpg_df3
mpg_df4
```


### mutate
```{r}
# get a ratio of hwy, cty
mutate(mpg_df4, ratio = hwy/cty)
```


### merge by columns
**Exercise** Try to use `inner_join()`, `full_join()`, `left_join()`, `right_join()` to merge `mpg_df3` and `mpg_df4` by columns



## Reshape data
convert the data to long/wide format using function `pivot_longer()` and `pivot_wider()`in tidyr
```{r}
mpg_df4_long = pivot_longer(mpg_df4, cols = c("hwy", "cty"), names_to = "type", values_to = "mile_per_gallon")
mpg_df4_long
```

Exercise, try to convert the `mpg_df4_long` to the wide format


## Group & Summarize 
group by manufacturer and year, and compute the mean of `hwy` and `cty`
```{r}
mpg_group = group_by(mpg, year, manufacturer)
mpg_summary = summarise(mpg_group, hwy_mean = mean(hwy), cty_mean = mean(cty))
mpg_summary
```




## Pipe
use pipe to chain the functions together
```{r}
mpg %>% group_by(year, manufacturer) %>% 
  summarise(hwy_mean = mean(hwy), cty_mean = mean(cty))
```

**Exercise** select the observation that has the smallest `cty` for each year.



# Aesthetic mapping
```{r}
base = ggplot(data = mpg)
base + geom_point(aes(x = displ, y = hwy, color = drv))
base + geom_point(aes(x = displ, y = hwy, size  = drv))
base + geom_point(aes(x = displ, y = hwy, alpha = drv))
base + geom_point(aes(x = displ, y = hwy, shape = drv))
```

## add, override, or remove mappings
```{r}
base = ggplot(mpg, aes(displ, hwy, colour =class))
base + geom_point()

## add mapping
base + geom_point(aes(shape  = drv))

## override mapping
base + geom_point(aes(colour = drv))

## remove mapping
base + geom_point(aes(colour = NULL))
```

## transformaiton in mapping
```{r}
ggplot(mpg) + geom_point(aes(x = displ, y = hwy, colour = class))
ggplot(mpg) + geom_point(aes(x = log(displ), y = hwy, colour = class))
```


## Aesthetic mapping vs setting
```{r}
ggplot(mpg) + geom_point(aes(displ, hwy, colour = "blue"))
ggplot(mpg) + geom_point(aes(displ, hwy), colour = "blue")
```

**Exercise**: build a plot that contains the following elements
- set `hwy` as x axis, and `cty` as `y` axis
- set point shapes and color according to `cyl`
- set theme as `theme_bw()`
- add title `scatter plot` and subtitle `cty.vs.hwy`

> Hint: the variable types matter!


