---
title: 'Day3: Examples and useful packages'
author: "Wenbin Guo"
date: "2023-11-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Beginning remarks
## goals
For today's workshop session, we'll try the following tasks
- customize the figures with legends and themes
- 2 complicated examples 
- try some useful packages

# load library
```{r}
library(tidyverse)

data(mpg)
head(mpg)
```

# customize the legend and figure
```{r}
base = ggplot(data = mpg) + geom_point(aes(x = displ, y = hwy, color = drv)) 
base
```

## legend
change legend label and title
```{r}
base + 
  scale_color_discrete(name = "Type", labels = c("4 wheel", "front drive", "rear drive"))
```


Exercise: 
change the dot color to `red`, `blue`, `green`, rename the axis label to the full name `displacement` and `highway`, add title and subtitle to the plots


## theme
```{r}
base + theme_bw()
base + theme_classic()
base + theme_dark()
base + theme_grey()
base + theme_light()
base + theme_linedraw()
base + theme_minimal()
base + theme_test()
base + theme_void()
```

Exercise:
- increase the size of the legend key and labels
- save the figure to a file called `plot.pdf`


# practical examples
## scatter plot
try to reproduce the `scatter_plot.png` figure using dataset `midwest`

```{r}
# data("midwest")
# 
# ggplot(midwest, aes(x=area, y=poptotal)) + 
#   geom_point(aes(col=state, size=popdensity)) + 
#   geom_smooth(method="loess", se=F) + 
#   xlim(c(0, 0.1)) + 
#   ylim(c(0, 500000)) + 
#   labs(subtitle="Area Vs Population", 
#        y="Population", 
#        x="Area", 
#        title="Scatterplot", 
#        caption = "Source: midwest") +
#   theme_bw()
```


## boxplot + dotplot
try to reproduce the `box_dotplot.png` figure using dataset `mpg`

```{r}
# g <- ggplot(mpg, aes(manufacturer, cty))
# g + geom_boxplot() + 
#   geom_dotplot(binaxis='y', 
#                stackdir='center', 
#                fill="red") +
#   theme(axis.text.x = element_text(angle=45, vjust=0.5)) + 
#   labs(title="Box plot + Dot plot", 
#        subtitle="City Mileage vs Class: Each dot represents 1 row in source data",
#        caption="Source: mpg",
#        x="Class of Vehicle",
#        y="City Mileage") +
#   theme_bw()
```


# useful packages
## esquisse
[esquisse](https://cran.r-project.org/web/packages/esquisse/vignettes/get-started.html)
```{r}
library(esquisse)

# demo
```


## ggpubr
[ggpubr](https://rpkgs.datanovia.com/ggpubr/)
```{r}
library(ggpubr)
my_comparisons <- list( c("f", "r"), c("r", "4"), c("f", "4") )

ggviolin(mpg, x = "drv", y = "hwy", fill = "drv",
         palette = c("#00AFBB", "#E7B800", "#FC4E07"),
         add = "boxplot", add.params = list(fill = "white"))+
  stat_compare_means(comparisons = my_comparisons, label = "p.signif")+ # Add significance levels
  stat_compare_means(label.y = 50)
```



## ggExtra
[ggExtra](https://cran.r-project.org/web/packages/ggExtra/vignettes/ggExtra.html)
```{r}
library(ggExtra)

g <- ggplot(mpg, aes(cty, hwy)) + 
  geom_point() + 
  geom_smooth(method="lm", se=F)

g
ggMarginal(g, type = "histogram", fill="transparent")
ggMarginal(g, type = "boxplot", fill="transparent")
ggMarginal(g, type = "density", fill="transparent")
```

## patchwork
[patchwork](https://patchwork.data-imaginist.com/)

```{r}
library(patchwork)

p1 = ggplot(mpg, aes(displ, cty)) + 
  geom_point() + 
  geom_smooth(method="lm", se=F)

p2 = ggplot(mpg, aes(displ, hwy)) + 
  geom_point() + 
  geom_smooth(method="lm", se=F)

p3 = ggplot(mpg, aes(cty, hwy)) + 
  geom_point() + 
  geom_smooth(method="lm", se=F)


# patch together
p1+p2

# patch 
(p1 | p2)/p3
```


- ggcorrplot: https://cran.r-project.org/web/packages/ggcorrplot/readme/README.html
- ggridge: https://cran.r-project.org/web/packages/ggridges/vignettes/introduction.html
- ggbreak: https://cran.r-project.org/web/packages/ggbreak/vignettes/ggbreak.html
- ggsci: https://nanx.me/ggsci/


see a full list of extensions: https://exts.ggplot2.tidyverse.org/gallery/
