---
title: 'Day2: Building a plot layer by layer'
author: "Wenbin Guo"
date: "2023-11-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Beginning remarks
## goals
For today's workshop session, we'll engage in exercises designed to deepen our understanding and proficiency in the following parts:
- Exploring aesthetic mappings 
- Delving into geometric concepts to strengthen the foundation of our design approach.
- Learning to use of scale to set customized color
- Utilizing facets effectively to display data
- Navigating the coordinate system

These practices aim to refine our skills and enrich our overall learning experience.

# load library
```{r}
library(tidyverse)

data(mpg)
head(mpg)
```

The meaning of the variables
- `cty` and `hwy` record miles per gallon (mpg) for city and highway driving.
- `displ` is the engine displacement in litres.
- `drv` is the drivetrain: front wheel (f), rear wheel (r) or four wheel (4).
- `model` is the model of car. There are 38 models.
- `class` is a categorical variable describing the type of car: two seater, SUV, compact, etc.


# aesthetics
## aesthetics mapping
```{r}
base = ggplot(data = mpg)
base + geom_point(aes(x = displ, y = hwy, color = drv))
base + geom_point(aes(x = displ, y = hwy, size  = drv))
base + geom_point(aes(x = displ, y = hwy, alpha = drv))
base + geom_point(aes(x = displ, y = hwy, shape = drv))
```



## aesthetics mapping vs setting
```{r}
base + geom_point(aes(x = displ, y = hwy, color = "blue"))
base + geom_point(aes(x = displ, y = hwy),color = "blue")
```


# geometry
## single variable
### bar plot
```{r}
base + geom_bar(aes(x = class))
```

Exercise 1: plot the barplot in ascending order (bar height increases from left to right)
Exercise 2: plot the barplot for each class where bar height is the mean of hwy of each group



### histogram
```{r}
base + geom_histogram(aes(x = hwy))
base + geom_histogram(aes(x = hwy), binwidth = 5)
```


### freqpoly
```{r}
base + geom_freqpoly(aes(x = hwy))
```


### density
```{r}
base + geom_density(aes(x = hwy))
```


### dotplot
```{r}
base + geom_dotplot(aes(x = hwy))
```

## two variables
### scatter plot
```{r}
base + geom_point(aes(x = displ, y =hwy))
```

### smooth 
```{r}
base + 
  geom_point(aes(x = displ, y =hwy)) + 
  geom_smooth(aes(x = displ, y =hwy), method = "lm")

base + 
  geom_point(aes(x = displ, y =hwy)) + 
  geom_smooth(aes(x = displ, y =hwy), method = "loess")

base + 
  geom_point(aes(x = displ, y =hwy)) + 
  geom_smooth(aes(x = displ, y =hwy), method = "loess") +
  geom_rug(aes(x = displ, y =hwy))
```

### barplot
```{r}
base + 
  geom_bar(aes(x = drv, y =hwy), stat = "identity")
```

### boxplot
```{r}
ggplot(mpg) +
  geom_boxplot(aes(x = drv, y = hwy)) 
```

### violin plot
```{r}
ggplot(mpg) +
  geom_violin(aes(x = drv, y = hwy))
```

### jitter
```{r}
ggplot(mpg) +
  geom_point(aes(x = drv, y = hwy)) +
  geom_boxplot(aes(x = drv, y = hwy))

ggplot(mpg) +
  geom_point(aes(x = drv, y = hwy), position = "jitter") +
  geom_boxplot(aes(x = drv, y = hwy))
```

### 2d density
```{r}
ggplot(mpg) +
  geom_hex(aes(x = displ, y = hwy)) 
```


### error bar and point range
```{r}
y <- c(18, 11, 16)
df <- data.frame(x = 1:3, y = y, se = c(1.2, 0.5, 1.0))

base <- ggplot(df, aes(x, y, ymin = y - se, ymax = y + se))
base + geom_crossbar()
base + geom_pointrange()
```


### heatmap (using geom_tile)
```{r}
ht_data  = expand.grid(X = 1:10, Y = 1:10)
ht_data$Value = runif(100, min = 1, max = 100)

ggplot(ht_data) + 
  geom_tile(aes(x = X, y = Y, fill = Value))
```

Note: this is just a demo of using ggplot2 to generate heatmap. Package `pheatmap`, `complexHeatmap`, `ggheatmap` might be more suitable for this task


# scale
Exercise: make the following boxplot to have color "red", "blue", "green"
```{r}
ggplot(mpg) +
  geom_point(aes(x = drv, y = hwy), position = "jitter") +
  geom_boxplot(aes(x = drv, y = hwy))
```


# facet
```{r}
base = ggplot(data = mpg)
base + geom_point(aes(displ, hwy, colour = drv)) + facet_null()

base + geom_point(aes(displ, hwy, colour = drv)) + facet_wrap(~class)

base + geom_point(aes(displ, hwy, colour = drv)) + facet_wrap(~class, scales = "free_x")

base + geom_point(aes(displ, hwy, colour = drv)) + facet_grid(year~class, scales = "free")
```

### facet with missing
```{r}
df1 <- data.frame(x = 1:3, y = 1:3, gender = c("f", "f", "m"))
df2 <- data.frame(x = 2, y = 2)

ggplot(df1, aes(x, y)) + 
  geom_point(data = df2, colour = "red", size = 2) + 
  geom_point() + 
  facet_wrap(~gender)
```



# coordinates
## linear
### zoom in
```{r}
base <- ggplot(mpg, aes(displ, hwy)) + 
  geom_point() + 
  geom_smooth()
```

```{r}
base + scale_x_continuous(limits = c(4, 6))
```

```{r}
base + coord_cartesian(xlim = c(4, 6))
```


### flipping axes
```{r}
ggplot(mpg, aes(displ, cty)) + 
  geom_point() + 
  geom_smooth()

ggplot(mpg, aes(cty, displ)) + 
  geom_point() + 
  geom_smooth()

ggplot(mpg, aes(displ, cty)) + 
  geom_point() + 
  geom_smooth() + 
  coord_flip()
```

### fixed ratio
```{r}
ggplot(mpg, aes(displ, cty)) + 
  geom_point() + 
  geom_smooth() 

ggplot(mpg, aes(displ, cty)) + 
  geom_point() + 
  geom_smooth() +
  coord_fixed(ratio = 0.2)
```



## nonlinear
### transofmration
```{r}
base <- ggplot(diamonds, aes(carat, price)) + 
  stat_bin2d() + 
  geom_smooth(method = "lm") + 
  xlab(NULL) + 
  ylab(NULL) + 
  theme(legend.position = "none")
base

# Better fit on log scale, but harder to interpret
base +
  scale_x_log10() + 
  scale_y_log10()

# Fit on log scale, then backtransform to original.
# Highlights lack of expensive diamonds with large carats
pow10 <- scales::exp_trans(10)
base +
  scale_x_log10() + 
  scale_y_log10() + 
  coord_trans(x = pow10, y = pow10)
```

### polar coordinates
```{r}
base <- ggplot(mtcars, aes(factor(1), fill = factor(cyl))) +
  geom_bar(width = 1) + 
  theme(legend.position = "none") + 
  scale_x_discrete(NULL, expand = c(0, 0)) +
  scale_y_continuous(NULL, expand = c(0, 0))

# Stacked barchart
base

# Pie chart
base + coord_polar(theta = "y")

# The bullseye chart
base + coord_polar()
```

