---
title: 'Day2: Building a plot layer by layer'
author: "Wenbin Guo"
date: "2023-02-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Beginning remarks
## Day2 outline
For today's workshop practice, we will cover the following topics
- Exploring geometric objects 
- Learning to use of scale to set customized color
- Utilizing facets effectively to display data
- Navigating the coordinate system


# load library
```{r}
library(tidyverse)

data(mpg)
```

# layers
```{r}
base = ggplot(mpg) 

base
base + aes(displ, hwy)

base + layer(mapping = aes(displ, hwy), geom = "point", stat = "identity", position = "identity")
base + geom_point(aes(displ, hwy))
```


# geometry
## basic types
### blank
```{r}
base + geom_blank()
```


### lines
```{r}
df = data.frame(x1 = c(1), y1 = c(20), x2 = c(3), y2 = c(30))
df2= data.frame(x = c(1,2,2.5,3,3.5,4), y = c(18,22, 36, 25, 20, 40))

# vline, hline, abline
base + aes(displ, hwy) + 
  geom_vline(xintercept = 3) +
  geom_hline(yintercept = 25) 

base + aes(displ, hwy) + 
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "blue") 



# curve + segment
base + aes(displ, hwy) + 
  geom_curve(data = df, aes(x = x1, y = y1, xend = x2, yend = y2, colour = "curve"), 
             curvature = 0.2, arrow = arrow(length = unit(0.03, "npc"))) +
  geom_segment(data = df, aes(x = x1, y = y1, xend = x2, yend = y2, colour = "segment"))


# line vs path 
base + aes(displ, hwy) + 
  geom_point() +
  geom_line()

base + aes(displ, hwy) + 
  geom_path()
```

**Exercise**: for `geom_curve` how to control the curved direction? (e.g. convex -> concave)

**Exercise**: plot the barplot for each class where bar height is the mean of hwy of each group



### region
```{r}
base + aes(displ, hwy) + 
  geom_rect(data = df, mapping = aes(xmin = x1, xmax = x2, ymin = y1, ymax = y2), 
            alpha = 0.5, inherit.aes = FALSE)

base + aes(displ, hwy) + 
  geom_polygon(data = df2, mapping = aes(x =x, y = y), inherit.aes = FALSE)
```


## single variable
### bar plot
```{r}
base + geom_bar(aes(x = class))
```

**Exercise**: plot the barplot in ascending order (bar height increases from left to right)


### histogram
```{r}
base + geom_histogram(aes(x = hwy))
base + geom_histogram(aes(x = hwy), binwidth = 5)
```


### freqpoly
```{r}
base + geom_freqpoly(aes(x = hwy))
```


### density
```{r}
base + geom_density(aes(x = hwy))
```


### dotplot
```{r}
base + geom_dotplot(aes(x = hwy))
```


## two variables
### scatter plot
```{r}
base + geom_point(aes(x = displ, y =hwy))
```

### smooth 
```{r}
base + 
  geom_point(aes(x = displ, y =hwy)) + 
  geom_smooth(aes(x = displ, y =hwy), method = "lm")

base + 
  geom_point(aes(x = displ, y =hwy)) + 
  geom_smooth(aes(x = displ, y =hwy), method = "loess")

base + 
  geom_point(aes(x = displ, y =hwy)) + 
  geom_smooth(aes(x = displ, y =hwy), method = "loess") +
  geom_rug(aes(x = displ, y =hwy))
```

### barplot
```{r}
base + 
  geom_bar(aes(x = drv, y =hwy), stat = "identity")
```

### boxplot
```{r}
ggplot(mpg) +
  geom_boxplot(aes(x = drv, y = hwy)) 
```

### violin plot
```{r}
ggplot(mpg) +
  geom_violin(aes(x = drv, y = hwy))
```

### jitter
```{r}
ggplot(mpg) +
  geom_point(aes(x = drv, y = hwy)) +
  geom_boxplot(aes(x = drv, y = hwy))

ggplot(mpg) +
  geom_point(aes(x = drv, y = hwy), position = "jitter") +
  geom_boxplot(aes(x = drv, y = hwy))
```

### 2d density
```{r}
ggplot(mpg) +
  geom_hex(aes(x = displ, y = hwy)) 
```


### error bar and point range
```{r}
y <- c(18, 11, 16)
df <- data.frame(x = 1:3, y = y, se = c(1.2, 0.5, 1.0))

p <- ggplot(df, aes(x, y, ymin = y - se, ymax = y + se))
p + geom_crossbar()
p + geom_pointrange()
```
## three variables
### contour 
```{r}
data(faithfuld)

ggplot(faithfuld, aes(waiting, eruptions, z = density)) +
  geom_contour()

ggplot(faithfuld, aes(waiting, eruptions, z = density)) +
  geom_contour_filled()
```


### heatmap (using geom_tile)
```{r}
ht_data  = expand.grid(X = 1:10, Y = 1:10)
ht_data$Value = runif(100, min = 1, max = 100)

ggplot(ht_data) + 
  geom_tile(aes(x = X, y = Y, fill = Value))
```

Note: There are more specialized packages such as `pheatmap`, `complexHeatmap`, `ggheatmap` that can be used to make fancy heatmaps


# position adjustment
## points
```{r}
base + geom_point(aes(drv, hwy))
```

```{r}
base + geom_point(aes(drv, hwy), position = "jitter")
base + geom_point(aes(drv, hwy), position = position_jitter())
base + geom_point(aes(drv, hwy), position = position_jitter(seed = 42))
```

## bar plot
```{r}
# stack
base + geom_bar(aes(manufacturer, hwy, fill = drv), position = position_stack(), stat = "identity")

# fill 
base + geom_bar(aes(manufacturer, hwy, fill = drv), position = position_fill(), stat = "identity")

# dodge
base + geom_bar(aes(manufacturer, hwy, fill = drv), position = position_dodge(), stat = "identity")
```

**Exercise**: change the position adjustment to be identity, what will you find?


# scale
**Exercise**: make the following boxplot to have color "red", "blue", "green"
```{r}
ggplot(mpg) +
  geom_point(aes(x = drv, y = hwy), position = "jitter") +
  geom_boxplot(aes(x = drv, y = hwy))
```


# facet
```{r}
base = ggplot(data = mpg)
base + geom_point(aes(displ, hwy, colour = drv)) + facet_null()

base + geom_point(aes(displ, hwy, colour = drv)) + facet_wrap(~class)

base + geom_point(aes(displ, hwy, colour = drv)) + facet_wrap(~class, scales = "free_x")

base + geom_point(aes(displ, hwy, colour = drv)) + facet_grid(year~class, scales = "free")
```

### facet with missing
```{r}
df1 <- data.frame(x = 1:3, y = 1:3, gender = c("f", "f", "m"))
df2 <- data.frame(x = 2, y = 2)

ggplot(df1, aes(x, y)) + 
  geom_point(data = df2, colour = "red", size = 2) + 
  geom_point() + 
  facet_wrap(~gender)
```


# coordinates
## linear
### zoom in
```{r}
base <- ggplot(mpg, aes(displ, hwy)) + 
  geom_point() + 
  geom_smooth()

base
```

```{r}
base + scale_x_continuous(limits = c(4, 6))
```

```{r}
base + coord_cartesian(xlim = c(4, 6))
```


### flipping axes
```{r}
ggplot(mpg, aes(displ, cty)) + 
  geom_point() + 
  geom_smooth()

ggplot(mpg, aes(cty, displ)) + 
  geom_point() + 
  geom_smooth()

ggplot(mpg, aes(displ, cty)) + 
  geom_point() + 
  geom_smooth() + 
  coord_flip()
```

### fixed ratio
```{r}
ggplot(mpg, aes(displ, cty)) + 
  geom_point() + 
  geom_smooth() 

ggplot(mpg, aes(displ, cty)) + 
  geom_point() + 
  geom_smooth() +
  coord_fixed(ratio = 0.2)
```



## nonlinear
### transofmration
```{r}
base <- ggplot(diamonds, aes(carat, price)) + 
  stat_bin2d() + 
  geom_smooth(method = "lm") + 
  xlab(NULL) + 
  ylab(NULL) + 
  theme(legend.position = "none")
base

# Better fit on log scale, but harder to interpret
base +
  scale_x_log10() + 
  scale_y_log10()

# Fit on log scale, then backtransform to original.
# Highlights lack of expensive diamonds with large carats
pow10 <- scales::exp_trans(10)
base +
  scale_x_log10() + 
  scale_y_log10() + 
  coord_trans(x = pow10, y = pow10)
```

### polar coordinates
```{r}
base <- ggplot(mtcars, aes(factor(1), fill = factor(cyl))) +
  geom_bar(width = 1) + 
  theme(legend.position = "none") + 
  scale_x_discrete(NULL, expand = c(0, 0)) +
  scale_y_continuous(NULL, expand = c(0, 0))

# Stacked barchart
base

# Pie chart
base + coord_polar(theta = "y")

# The bullseye chart
base + coord_polar()
```


# Valentine's special
## generate heart curve
**Exercise**: generate a heart curve using ggplot2 and assign the plot into a variable `heart`
```{r}

```

## save into image
```{r}
myheart_png = "./myheart.png"
ggsave(heart, filename = myheart_png)
```


## modify the image
let's add some arts to the creation
```{r}
if(!require(brickr)){
   remotes::install_github("ryantimpe/brickr")
}
if(!require(png)){
   install.packages("png")
}
if(!require(animation)){
   install.packages("animation")
}
library(brickr)
library(png)
library(animation)
```

```{r}
img_mosaic = readPNG(myheart_png) %>%
  image_to_mosaic() 
```

```{r}
img_mosaic %>%
  build_mosaic()
```

```{r}
build = img_mosaic %>%
  build_instructions(num_steps = 6)

build
```

## extract subplots
```{r}
s1 = build
s1$data = s1$data %>% dplyr::filter(Step == "Step 01") 
s1 = s1 + scale_x_continuous(limits=c(0,50)) + scale_y_continuous(limits=c(0,50))

s2 = build
s2$data = s2$data %>% dplyr::filter(Step == "Step 02") 
s2 = s2 + scale_x_continuous(limits=c(0,50)) + scale_y_continuous(limits=c(0,50))

s3 = build
s3$data = s3$data %>% dplyr::filter(Step == "Step 03") 
s3 = s3 + scale_x_continuous(limits=c(0,50)) + scale_y_continuous(limits=c(0,50))


s4 = build
s4$data = s4$data %>% dplyr::filter(Step == "Step 04") 
s4 = s4 + scale_x_continuous(limits=c(0,50)) + scale_y_continuous(limits=c(0,50))

s5 = build
s5$data = s5$data %>% dplyr::filter(Step == "Step 05") 
s5 = s5 + scale_x_continuous(limits=c(0,50)) + scale_y_continuous(limits=c(0,50))

s6 = build
s6$data = s6$data %>% dplyr::filter(Step == "Step 06") 
s6 = s6 + scale_x_continuous(limits=c(0,50)) + scale_y_continuous(limits=c(0,50))
```


## save into a gif
```{r}
animation::saveGIF(
  expr = {
    plot(s1)
    plot(s2)
    plot(s3)
    plot(s4)
    plot(s5)
    plot(s6)
  },
  movie.name = "explicit_my3.gif"
)
```
